\documentclass[../book/calcnotes.tex]{subfiles}

\begin{document}
\section{The fundamental theorem of integral calculus}
\label{sec:ftc}

In the last section, we defined the \emph{definite integral} $\Int{f(x)}{x,a,b}$ of a function $f$ over an interval $\closedint{a,b}$ to be the \enquote{signed area} of the region bounded between the curve $y = f(x)$ and the $x$-axis.
However, this doesn't really solve our original problem!
We want \emph{methods} for computing these areas, not just a name and some notation.

To progress towards this goal, we're going to need some heavy machinery---and, perhaps surprisingly, we'll need the tools of differential calculus!
This suggests a refined version of our motivating problem:
\begin{motprob}
  Given a function $f$ which is integrable over some interval $I = \sbrac{a, b}$, what is the analytic behavior of the \enquote{area function} $\areafunc{f}{a}{x} = \Int{f(t)}{t,a,x}$?
\end{motprob}

\subsection{Integrals and derivatives}
\label{sec:integrals-and-derivatives}

First, let's get our feet wet with a quick example involving this area function.
\begin{example}
  Let $f(x) = 2x+1$.
  What is the area function $\areafunc{f}{1}{x}$?
  What is its derivative?
\end{example}

\begin{soln}
  We first compute the area function $\areafunc{f}{1}{x}$.
  Since $f$ is linear, the region bounded by $y = f(x)$ over $\sbrac{1, x}$ is a trapezoid, as shown in \cref{fig:areafunc}
  The left height of this trapezoid is $1$, its right height is $2 \cdot x - 1$, and its base is $x - 1$.
  Thus, its area is $\areafunc{f}{1}{x} = \pbrac{x-1} \cdot \frac{1}{2} \pbrac{2x-1 + 1} = x^{2} - x$.
  It is then straightforward to compute that $\D{}{x} \areafunc{f}{1}{x} = 2x-1$; we note that this is in fact $f(x)$ again.

  \begin{smallfig}
    \begin{tikzpicture}
      \begin{axis}[
        agdplot,
        ymin = -1.2,
        ymax = 8,
        xmin = -1,
        xmax = 4,
        ytick = \empty,
        xtick = {1},
        extra x ticks = {2.5},
        extra x tick labels = {$x$},
        ]
        \addplot+ [domain=1:2.5,box1] {2*x-1} \closedcycle;

        \addplot [thick, domain=-1:4] {2*x-1} node [pos=0.45,pin={90:$y = 2x-1$}] {};

      \end{axis}
    \end{tikzpicture}
    \caption{Region bounded by $y = 2x-1$ over $\closedint{1,x}$}
    \label{fig:areafunc}
  \end{smallfig}
\end{soln}

Now \emph{that's} a strange coincidence.
Why on Earth should it be the case that $\D{}{x} \areafunc{f}{1}{x} = f(x)$?
Is this true for other functions?

In fact, it's actually true for \emph{all} integrable functions.
To see why, let's look at an example more closely.
Consider \cref{fig:ftc}.

\begin{widefig}
  \begin{tikzpicture}
    \begin{axis}[
      agdplot,
      rectify,
      ymin = -.2,
      ymax = .6,
      xmin = -.2,
      xmax = 5.5,
      ytick = \empty,
      xtick = \empty,
      extra x ticks = {1,4.5,4.75},
      extra x tick labels = {$a$,$x$,$x+\Delta x$},
      tick label style = {rotate = 45, anchor = east},
      ]
      \pgfmathdeclarefunction{ftcfunc}{1}{\pgfmathparse{1/3+1/4*sin(deg(pi*#1^2/8))}}

      \addplot+ [domain=1:4.5,box1] {ftcfunc(x)} \closedcycle;
      \addplot+ [domain=4.5:4.75,box2] {ftcfunc(x)} \closedcycle;

      \addplot [thick, domain=-1:8] {ftcfunc(x)} node [pos=0.45,pin={90:$y = f(x)$}] {};
    \end{axis}
  \end{tikzpicture}
  \caption{Demonstration of the fundamental theorem of calculus}
  \label{fig:ftc}
\end{widefig}

$\D{}{x} \areafunc{f}{a}{x}$ measures the rate of change of the area under $y = f(x)$ as the right border of the region is moved.
Let's use \cref{def:deriv} to unpack what this means.
Formally, we have that
\begin{align*}
  \D{}{x} \areafunc{f}{a}{x} &= \lim_{\Delta x \to 0} \frac{\areafunc{f}{a}{x + \Delta x} - \areafunc{f}{a}{x}}{\Delta x} \\
  &= \lim_{\Delta x \to 0} \frac{\Int{f(t)}{t, a, x + \Delta x} - \Int{f(t)}{t, a, x}}{\Delta x},
  \intertext{So, applying \cref{thm:int.combint}, we have that}
  &= \lim_{\Delta x \to 0} \frac{1}{\Delta x} \Int{f(t)}{t, x, x + \Delta x}.
\end{align*}

This integral measures the area under the curve of $y = f(x)$ over the interval $\closedint{x, x + \Delta x}$.
if $\Delta x$ is small enough, this is essentially a rectangle of height $f(x)$, whose area is of course given by $f(x) \cdot \Delta x$.
Thus, $\D{}{x} \areafunc{f}{a}{x} = f(x)$!

This result will turn out to be \emph{incredibly} important in what's to come, so we should record it as a theorem.
In fact, this result is so important that history has hallowed it with a very fancy name---no small feat, given the solemnity of your average mathematician!

\begin{theorem}[Fundamental theorem of integral calculus I]
  \label{thm:ftc.deriv}
  If $f$ is integrable over $\closedint{a,b}$ and $x \in \openint{a,b}$, then
  \begin{equation}
    \label{eq:ftc.deriv}
    \D{}{x} \Int{f(t)}{t,a,x} = f(x).
  \end{equation}
  In other words, the area function $\areafunc{f}{a}{x}$ is an antiderivative of $f$.
\end{theorem}

Taken at face value, this theorem doesn't seem all that interesting---after all, we've already seen in \cref{sec:antiderivs} how to find antiderivatives of functions, and integration is a lot of work!
Why have we gone to all this trouble just to find one more antiderivative of a function?

In fact, this theorem isn't about finding antiderivatives at all; instead, it's about computing definite integrals!
Recall from \cref{thm:antideriv.constant} that the antiderivatives of a continuous\sidefootnote{The functions we're interested in are almost always continuous, so this restriction won't cost us much.} function are all the same up to a constant.
In particular, if $f$ is continuous and $F$ is \emph{any} antiderivative of $f$, then there's some number $C$ such that $\Int{f(t)}{t,a,x} = F(x) + C$ for \emph{all} values of $x$!
If we can figure out that value $C$, we'll have an \emph{extremely} powerful tool for computing definite integrals at very low cost.

So what is this number $C$?
If we rearrange the equation above, we find that $C = \Int{f(t)}{t,a,x} - F(x)$ for every $x$.
To make this easy to calculate, let's take $x = a$; then the integral is $0$, showing that $C = - F(a)$.
(That wasn't so hard!)
In other words, $\Int{f(t)}{t,a,x} = F(x) - F(a)$.

Bringing all this together, we can wrap up our most powerful results about definite integrals in the following theorem, which represents the payoff of all our analytic work on definite integrals.

\begin{theorem}[Fundamental theorem of integral calculus II]
  \label{thm:ftc.calc}
  If $f$ is continuous and integrable on $\closedint{a,b}$ and $F$ is an antiderivative of $f$ on $\closedint{a,b}$ then
  \begin{equation}
    \label{eq:ftc.calc}
    \Int{f(x)}{x,a,b} = F(b) - F(a).
  \end{equation}
\end{theorem}

Thus, if we want to compute a definite integral of a continuous function $f$, we don't need to bother with the geometry of the region under the curve.
Instead, we need only \emph{evaluate some antiderivative of $f$ at two points!}
This is a vast improvement over the ad-hoc methods we used in the last section, since it means we can easily integrate functions with very complicated geometry.

\begin{example}
  \label{ex:ftc.calc}
  Let $f(x) = x^{2}$.
  What is $\Int{f(x)}{x,0,4}$?
\end{example}

\begin{soln}
  We first find that $F(x) = \frac{1}{3} x^{3}$ is an antiderivative of the given function $f(x) = x^{2}$.
  Thus, by \cref{thm:ftc.calc}, $\Int{x^{2}}{x, 0, 4} = F(4) - F(0) = \frac{1}{3} \pbrac{4}^{3} - \frac{1}{3} \pbrac{0}^{3} = \frac{64}{3}$.
\end{soln}

Before we move on, take a moment to think about how much harder the calculation in \cref{ex:ftc.calc} would have been without \cref{thm:ftc.calc}.
We'd have needed to find the area under the curve of $y = x^{2}$ using purely geometric methods.
This problem was solved by Archimedes, but it required a \emph{lot} of work and extreme cleverness---and we just dispensed with it in one line of calculations!

\subsection{Net change}
\label{sec:netchange}

\Cref{thm:ftc.calc} suggests an important interpretation of the area function.
The function $\Int{f(t)}{t,a,x}$ doesn't just measure the area of a region in the plane; it measures the \emph{accumulation} or \emph{net change} of a quantity whose rate of change is given by $f$.
This interpretation is quite natural in settings where rates of change have physical interpretations.

\begin{example}
  \label{ex:flowint}
  Water is draining out of a tank through a pipe; at time $t$ (in \si{\minute}), its flow rate is $r(t)$ (in \si{\gallon\per\minute}).
  The definite integral $\Int{r(t)}{t,a,b}$ measures the total volume of water (in \si{\gallon}) that flows through the pipe during the time interval $t \in \closedint{a, b}$.
\end{example}

\begin{example}
  \label{ex:velocityint}
  A car is traveling on a highway from Minneapolis to Chicago; at time $t$ (in \si{\hour}), its velocity (in \si{\mile\per\hour}) is given by $v(t)$.
  The definite integral $\Int{f(t)}{t,a,b}$ measures the \emph{total displacement} of the car (in \si{\mile}) over the time interval $t \in \closedint{a,b}$; in particular, if $v$ is nonnegative, then the integral measures the distance traveled by the car, which is the distance between the car's positions at time $a$ and time $b$.
  (If the car changes direction during its trip, this is not the same as total distance traveled; it is important to bear in mind the significance of sign in this calculation!)
\end{example}

\exercises
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../book/calcnotes.tex"
%%% End: